---
title: "COVID-19 analysis based on crowdsourced data"
---

[Data source](https://docs.google.com/spreadsheets/d/e/2PACX-1vQU0SIALScXx8VXDX7yKNKWWPKE1YjFlWc6VTEVSN45CklWWf-uWmprQIyLtoPDA18tX9cFDr-aQ9S6/pubhtml)

Paper:
[Early epidemiological analysis of the coronavirus disease 2019 outbreak based on crowdsourced data: a population-level observational study](https://www.thelancet.com/journals/landig/article/PIIS2589-7500(20)30026-1/fulltext)

Date downloaded: 2020-03-11.
Latest data update: 2020-03-09 1PM EST.

DISCLAIMER:
Data quality is **very questionable**. As such, any conclusions will be very questionable too.
This is meant as a didactic exercise only. Students are invited to find better data and/or 
keep a very critical mindset while going through the results.

```{r}
library(tidyverse)
```

# Data preparation

```{r}
d_raw <- read_csv("covid19.csv",
              col_types = cols(id = 'c', case_in = 'c', age = 'd', if_onset_approximated = 'l',
                               international_traveler = 'l', domestic_traveler = 'l', traveler = 'l',
                               `visiting Wuhan` = 'l', `from Wuhan` = 'l',
                               .default = 'c'
                               ))
```

```{r}
#?
#d_raw <- d_raw[-2224,]
```

```{r}
table(d_raw$death)
```

```{r}
as_date <- function(x) as.Date(x, format = "%m/%d/%y")
d <-
  d_raw %>%
  mutate(reporting_date = as_date(reporting_date),
         hosp_visit_date = as_date(hosp_visit_date),
         exposure_start = as_date(exposure_start),
         exposure_end = as_date(exposure_end),
         symptom_onset = as_date(symptom_onset),
         death_status = death != "0",
         death_date = as.Date(ifelse(!death %in% c("0", "1"), as.Date(death, format = "%m/%d/%y", origin = "1970-01-01"), NA), origin = "1970-01-01"),
         gender = factor(gender, levels = c("female", "male")))
```

# Binary outcome: alive/dead

## Sex impact

### Frequency tables and independence tests

(1) Chi-Square: p-value > alpha = 0.05, cannot reject null hypothesis, gender and death are independent
```{r}
with(d, table(death_status, gender))
with(d, prop.table(table(death_status, gender), 2))
with(d, 100 * prop.table(table(death_status, gender), 2))
with(d, round(100 * prop.table(table(death_status, gender), 2), 1))
with(d, chisq.test(table(death_status, gender)))
```

(2) Fisher Test: p-value > alpha = 0.05, cannot reject null hypothesis, gender and death are independent
```{r}
with(d, fisher.test(table(death_status, gender)))
```

### Logistic regression
```{r}
summary(lm(death_status ~ gender, data = d))
```

```{r}
summary(glm(death_status ~ gender, data = d, family = "binomial"))
```

```{r}
exp(confint(glm(death_status ~ gender, data = d, family = "binomial")))
exp(confint(glm(death_status ~ gender, data = d, family = "binomial"))[2,])
confint(glm(death_status ~ gender, data = d, family = "binomial"))
```

```{r}
exp(0.5069)
```


## Age
```{r}
glm(death_status ~ age, data = d, family = "binomial")
```

```{r}
summary(glm(death_status ~ age, data = d, family = "binomial"))
```


```{r}
fit <- glm(death_status ~ I(age/10), data = d, family = "binomial")
summary(fit)
```

```{r}
exp(confint(fit)[2,])
```

```{r}
exp(coef(fit)[2])
```

## Smoothing splines for age

```{r}
library(mgcv)
```

```{r}
fit <- gam(death_status ~ s(age), data = d, family = "binomial")
```

```{r}
summary(fit)
```

```{r}
plot(fit)
```


### Quadratic term for age

```{r}
fit <- glm(death_status ~ age + I(age^2), data = d, family = "binomial")
summary(fit)
```

### Piece-wise linear terms

```{r}
d2 <- mutate(d, age_70 = ifelse(age > 70, age - 70, 0))
```

```{r}
fit <- glm(death_status ~ age + age_70, data = d2, family = "binomial")
summary(fit)
```

From 0 to 70:
```{r}
exp(coef(fit)[2])
```

After 70:
```{r}
exp(coef(fit)[2:3])
exp(sum(coef(fit)[2:3]))
```

### A different parametrization

```{r}
d3 <- mutate(d,
             age_l70 = ifelse(age <= 70, age, 70),
             age_g70 = ifelse(age > 70, age - 70, 0))
fit <- glm(death_status ~ age_l70 + age_g70, data = d3, family = "binomial")

summary(fit)
```

```{r}
exp(confint(fit)[2:3,])
```

## Comparing countries

Caveats:
1. obsolete data
2. different testing guidelines per country
3. different reporting accuracy (i.e., often only deaths from hospitals are counted)
4. ...?

```{r}
table(d$country)
```

```{r}
d1 <- mutate(d, country = relevel(factor(country), ref = "China"))
summary(glm(death_status ~ country, data = d1, family = "binomial"))
```

## Adjusted model

```{r}
fit <- glm(death_status ~ I(age/10) + gender + country, data = d1, family = "binomial")
summary(fit)
```

```{r}
exp(coef(fit)[2:3])
```

# Maximum Likelihood Estimation

```{r}
library(maxLik)
```

Model: death by gender
```{r}
dx <- d %>% select(gender, death_status) %>% na.omit()

x <- (dx$gender == "male") + 0
y <- dx$death_status + 0

logLik <- function(beta) {
  linear_predictor <- beta[1] + beta[2] * x
  log_probabilities <-
    dbinom(y, size = 1, prob = plogis(linear_predictor), log = TRUE)
  log_likelihood <- sum(log_probabilities)
  return(log_likelihood)
}
```

```{r}
debugonce(logLik)
```

Check for the accuracy:
```{r}
logLik(c(0, 0))
```

```{r}
fit <- maxLik(logLik, start = c(intercept = 0, gender = 0))
```

```{r}
summary(fit)
```

```{r}
summary(glm(death_status ~ gender, data = dx, family = "binomial"))
```

# Continuous outcome: survival time

```{r}
library(survival)
END_OF_STUDY <- as.Date("2020-03-10", format = "%Y-%m-%d", origin = "1970-01-01")
```

From first symptom to date of data collection: March 10th, 2020
```{r}
d_surv <-
  d %>%
  filter(!is.na(symptom_onset), !is.na(death_status),
         !(is.na(death_date) & death_status)) %>%
  mutate(
    death_date = as.Date(ifelse(is.na(death_date), END_OF_STUDY, death_date), origin = "1970-01-01"),
    time = difftime(death_date, symptom_onset, units = "days")
  )
```

```{r}
fit <- survfit(Surv(time, death_status) ~ 1, data = d_surv)
plot(fit)
```

Let's zoom in:
```{r}
plot(fit, ylim = c(0.95, 1),
     xlab = "time (days)", ylab = "Survival(t)")
```

Cannot extract a median, 7 deaths out of 623 cases (~1.1%):
```{r}
fit
```

## Effects of covariates on risk of death

```{r}
summary(d_surv)
```

## Gender

```{r}
summary(coxph(Surv(time, death_status) ~ gender, data = d_surv))
```

## Country

Model does not converge: again, too few cases.
```{r}
summary(coxph(Surv(time, death_status) ~ country, data = d_surv))
```
Is risk in France really 1.8 **billion** times that in China?

## Age

This one is strong, as expected:
```{r}
summary(coxph(Surv(time, death_status) ~ I(age / 10), data = d_surv))
```

Is the effect linear, though?
```{r}
fit_age <- coxph(Surv(time, death_status) ~ pspline(age, df = 4), data = d_surv)
summary(fit_age)
```

```{r}
termplot(fit_age, col.term = 1, col.se = 1)
```

## Piecewise-linear age effect

Segments: constant 0-70; increasing after 70

```{r}
e <-
  d_surv %>%
  mutate(
    age_70p = ifelse(age <= 70, 0, age - 70)
  )
```

```{r}
fit_age_segments <- coxph(Surv(time, death_status) ~ age_70p, data = e)
summary(fit_age_segments)
```

Or more simply, before and after 70:
```{r}
fit_age_binary <- coxph(Surv(time, death_status) ~ I(age > 70), data = e)
summary(fit_age_binary)
```

Risk after 70yo is **40 times** that of people less than 70yo!
Let's see it visually:
```{r}
e$x <- factor(ifelse(e$age > 70, ">70", "<=70"), levels = c("<=70", ">70"))
plot(survfit(Surv(time, death_status) ~ I(age > 70), data = e),
     ylim = c(0.8, 1),
     col = 1:2)
```

```{r}
dx <- d%>% select(gender, death_status) %>% na.omit()

x <- (dx$gender == "male") +0
y <- dx$death_status +0

logLik <- function(beta) {
  linear_predictor = beta[1] +beta[2]
  
}
```
